### 2021-04-19 paper/appendix_STA561_project.tex
GetPossiblePatterns function

\setstretch{1}
\begin{flushright} 
    \end{flushright}
    \begin{center} \textbf{Parlr: Parallelized Record Linkage in R}

@@

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Most methods for conducting record linkage, the task of identifying
duplicate records across datasets, are derived from the foundational
article by Fellegi and Sunter in 1969. Within this family of methods,
two methods have notable strengths: \texttt{fastlink}, proposed by
Enamorado et al (2019), continues a is able to conduct linkage on
incredibly large datasets

@@

Our notation and assumptions closely follow that of Sadinle (2017).
Denote two files as \(A\) and \(B\), with \(n_A\) and \(n_B\) records
respectively, and with records indexed as \(i \in \{1, \ldots, n_A\}\)
in \(A\) and \(j \in \{1, \ldots, n_B\}\) in \(B\). Without loss of
generality, label the files such that \(n_A \geq n_B\). We also assume
there are no duplicates within files, only across. For each record pair
under consideration, we generate a comparison vector
\(\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}\),
where \(F\) is the number of fields used in the linkage and each
\(\gamma_{ij}^f\) takes on a value \(l \in \{1, \ldots, L_f\}\)
indicating level agreement between the two records on a specified field.

@@

Wherever possible, we reserve superscripts for denoting field and level,
while reserving subscripts for record indices. For example,
\(\mathbf{m}^f = (m^{f1}, \ldots, m^{fL_f})\) is the probability
distribution governing field \(f\) for matching records, and
\(\mathbf{m}_{ij}= \prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(m^{fl}\right)^{\mathbf{1}_{\gamma_{ij}^f = l}} = P(\boldsymbol{\gamma}_{ij}|Z_j = i)\)
is product of the relevant of the appropriate \(\mathbf{m}\) parameters
for record pair \((i,j)\). We hope that these conventions avoid
overloaded notation in the likelihood and subsequent derivations.

@@

(This will be revised to talk more explicitly about sampling \(Z_j\)
independently)

Our model differs from that of Sadinle 2017 through its explicit
dependence on a beta random variable \(\lambda\) that models the rate of
matching across records. Sadinle marginalizes over such a random
variable in his derivations of the ``beta prior for bipartite
matching,'' but here we provide derivations without marginalizing in
order to specify differing rates of matching for different linkage
clusters.

\hypertarget{prior-distributions-and-likelihood}{%
\subsection{Prior Distributions and
Likelihood}\label{prior-distributions-and-likelihood}}

@@

@@

Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let \(H\)
be the set of unique agreement patterns in the data, let \(P\) denote
the total number of unique agreement patterns. Note that \(P\) is
bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not
scale with \(n_A\) or \(n_B\). We index these agreement patterns by
\(p \in \{1, \ldots, P\}\), and say \((i,j) \in h_p\) when the \((i,j)\)
pair exhibits the \(p^{th}\) agreement pattern. Wherever possible, we
conduct calculations over these \(P\) agreement patterns rather than the
\(n_A \times n_B\) record pairs.

@@

We store this information in a nested list \(\tilde{\Gamma}\) where the
\(p^{th}\) component of the \(j^{th}\) list contains a vector of records
in \(A\) that share agreement pattern \(h_p\) with record \(j \in B\).
For each \(p\), we also calculate
\(|h_p| = \sum_{i=1}^{n_A}\sum_{j=1}^{n_B} \mathbf{1}_{(i,j) \in h_p}\)
total instances of agreement pattern \(h_p\) throughout the data, and
also for each \(j\), we calculate
\(|h_p|_j = \sum_{i=1}^{n_A} \mathbf{1}_{(i,j) \in h_p}\) the instances
of agreement pattern \(p\) among the comparison vectors between record
\(j \in B\) and each of the \(n_A\) records in \(A\).

For large data, we can partition the two datasets \(A\) and \(B\) into
smaller blocks \(\{A_m\}\) and \$\{B\_m\} for more manageable
computations. On a single machine, we can read-in data sequentially,
conduct hashing, collect results, and delete the original data from
memory before continuing with the next chunk of data. With multiple
cores or multiple machines, this can be done in parallel. Storing this
hashed information still becomes burdensome for large data, but this
hashing method greatly expands the capabilities of the Fellegi-Sunter
framework.

@@

After receiving matching statuses from \(\mathbf{Z}\), the Sadinle
method calculates \(\alpha_{fl}(\mathbf{Z})\) and
\(\beta_{fl}(\mathbf{Z})\) for each field and level. This constitutes
\(2 \times \sum L_f\) many summations over \(n_A \times n_B\)
quantities, and becomes computationally burdensome with large data. In
contrast, we recognize that each unique agreement pattern contributes to
the posterior \(\alpha(\mathbf{Z})\) and \(\beta(\mathbf{Z})\) vectors
in the same way. In fact, if we denote
\(|h_p^m| = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}\) to be the
number of matching record pairs with agreement pattern \(h_p\), then the
contribution of pairs of pattern \(h_p\) to the \(\alpha(\mathbf{Z})\)
vector is simply \(\mathbf{Z}(h_p) \times h_p\). Thus our posterior
update for \(\alpha\) is simply
\(\alpha(\mathbf{Z}) = \alpha_0 + \sum_{p=1}^P |h_p^m| \times h_p\).
Then, we can easily calculate \(|h_p^u|\), the number of nonmatching
record pairs of agreement pattern \(p\), by subtracting the number of
matching pairs from the total present in the data; that is
\(|h_p^u| = |h_p| - |h_p^m|\). From this, we can update our \(\beta\)
parameter through
\(\beta(\mathbf{Z}) = \beta_0 + \sum_{p=1}^P |h_p^u| \times h_p\). Note
that these constitute \(P\) many summations over \(n_B\) quantities, and
thus avoids the \(n_A \times n_B\) summation from the original method.

Sadinle uses a prior for \(\mathbf{Z}\) that induces the a full
conditional for \(Z_j\) that strictly enforces one-to-one matching.
Particular, this sampler removed previously matches records from the set
of candidate records when sampling \(Z_j\), creating a dependency that
makes the sampler \emph{inherently serial}.By weakening the one-to-one
requirement, our full conditional for \(Z\) does not depend on the rest
of the \(\mathbf{Z}_{-j}\) vector, and thus can be computed in parallel.
More importantly, since only the agreement pattern of \(Z_j\) is used
for calculations within the Gibbs sampler, and not the particular record
label, we can conduct this sampling only at the level of the unique
agreement patterns. This boosts computation time far greater than
parallelization.

To do this we calculate the Fellegi Sunter weight \(w_{h_p}\) for each
unique pattern, sample the agreement pattern between \(j\) and its
potential match, and then sample the record label uniformly among viable
records. More concretely, define \(H(Z_j)\) to be the agreement pattern
between \(j\) and its potential match, and say \(H(Z_j) = h_{P+1}\) when
\(Z_j = n_A + 1\). Then,

\[H\left(Z_j^{(s+1)}\right) | \mathbf{m}, \mathbf{u}, \mathbf{Z^{(s)}} \propto
\begin{cases} 
    w_{h_p}\times |h_p|_j  & h(z_j) \in H; \\
     n_A \frac{n_B - D + \beta_{\lambda}}{D + \alpha_{\lambda}} &  z_j  = n_A + 1 \\
\end{cases}\]

@@

\hypertarget{enforcing-bipartite-matching}{%
\subsection{Enforcing Bipartite
Matching}\label{enforcing-bipartite-matching}}

Add this

@@

\begin{itemize}
\item
  Replicating the Sadinle Study
\item
  \texttt{parlr} and \texttt{BRL} speed comparison
\end{itemize}

\hypertarget{data-analysis}{%
\section{Data Analysis}\label{data-analysis}}

\begin{itemize}
\tightlist
\item
  Necessary? El Salvador data? Or perhaps needs to be a bigger dataset?
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We note that this often not a reasonable assumption; for example if we
know that two records differ on gender, this does not give any
information on whether their year of births are the same, but it does
suggest that its more likely that their first names are different.

>>>>

\setstretch{1}
\begin{flushright} 
    \end{flushright}
    \begin{center} \textbf{Parlr: Parallelized Record Linkage in R} \linebreak
    \textbf{Technical Appendix}

@@

@@

Denote two files as \(A\) and \(B\), with \(n_A\) and \(n_B\) records
respectively, with records indexed as \(i \in \{1, \ldots, n_A\}\) in
\(A\) and \(j \in \{1, \ldots, n_B\}\) in \(B\). Without loss of
generality, label the files such that \(n_A \geq n_B\). We also assume
there are no duplicates within files, only across.

For each record pair under consideration, we generate a comparison
vector
\(\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}\),
where \(F\) is the number of fields used in the linkage and each
\(\gamma_{ij}^f\) takes on a value \(l \in \{1, \ldots, L_f\}\)
indicating the level agreement between the two records on a specified
field. For ease of modeling and computation, we restrict these
similarity scores to be discrete, ordinal variables, and the
construction of these is left to the modeler. With notation
\(\gamma_{ij}^f \in \{0, \ldots, L_f - 1\}\) for the number of possible
agreement levels for field \(f\), we use binary 0-1 variables to
indicate exact matching, and 0-1-2 variables to provide an option for
partial matching. For text data, we calculate similarity based on
Levenstein distance or some other text similarity score, and bin these
scores to integers for use in the model.

@@

Wherever possible, we reserve superscripts for denoting field and level,
while reserving subscripts for record indices. For example,
\(\mathbf{m}^f = (m^{f1}, \ldots, m^{fL_f})\) is the probability
distribution governing field \(f\) for matching records, and
\(\mathbf{m}_{ij}= \prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(m^{fl}\right)^{\mathbf{1}_{\gamma_{ij}^f = l}} = P(\boldsymbol{\gamma}_{ij}|Z_j = i)\)
is the product of the appropriate \(\mathbf{m}\) parameters for record
pair \((i,j)\). We hope that these conventions avoid overloaded notation
in the likelihood and subsequent derivations.

@@

@@

Since our Gibbs procedure does not strictly enforce one-to-one matching,
it is possible for the final Bayes estimate to link multiple records in
\(B\) to one record in \(A\). The modeler can either report both such
matches (with their respective posterior match probabilities), or
resolve these conflicts by accepting only the match with highest
posterior probability. Rigorous proof of this resolution procedure is
deferred to future work.

@@

Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P\) denote the total number of unique agreement patterns. Note that
\(P\) is bounded above by \(\prod_{f=1}^F L_f\), and that this bound
does not scale with \(n_A\) or \(n_B\). We index these agreement
patterns by \(p \in \{1, \ldots, P\}\), and say \((i,j) \in h_p\) when
the \((i,j)\) pair exhibits the \(p^{th}\) agreement pattern. Wherever
possible, we conduct calculations over these \(P\) agreement patterns
rather than the \(n_A \times n_B\) record pairs.

@@

We store this information in a nested list \(\tilde{\Gamma}\) where the
\(p^{th}\) component of the \(j^{th}\) list contains a vector of records
in \(A\) that share agreement pattern \(p\) with record \(j \in B\). For
each \(p\), we also calculate
\(H_p = \sum_{i=1}^{n_A}\sum_{j=1}^{n_B} \mathbf{1}_{(i,j) \in h_p}\),
the total instances of agreement pattern \(p\) throughout the data, and
also for each \(j\), we calculate
\(H_{p_j} = \sum_{i=1}^{n_A} \mathbf{1}_{{(i,j) \in h_p}}\) the
instances of agreement pattern \(p\) among the comparison vectors
between record \(j \in B\) and each of the \(n_A\) records in \(A\).

For large data, we can partition the two datasets \(A\) and \(B\) into
smaller blocks \(\{A_m\}\) and \(\{B_m\}\) for more manageable
computations. On a single machine, we can read-in data sequentially,
conduct hashing, collect results, and delete the original data from
memory before continuing with the next chunk of data. With multiple
cores or multiple machines, this can be done in parallel. Storing this
hashed information still becomes burdensome for large data, but this
hashing method greatly expands the capabilities of the Fellegi-Sunter
framework.

@@

After receiving matching statuses from \(\mathbf{Z}\), the Sadinle
method calculates \(\alpha_{fl}(\mathbf{Z})\) and
\(\beta_{fl}(\mathbf{Z})\) for each field and level. This constitutes
\(2 \times \sum L_f\) many summations over \(n_A \times n_B\)
quantities, and becomes computationally burdensome with large data. In
contrast, we recognize that each unique agreement pattern contributes to
the posterior \(\alpha(\mathbf{Z})\) and \(\beta(\mathbf{Z})\) vectors
in the same way. In fact, if we denote
\(H_p^m = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}\) to be the
number of matching record pairs with agreement pattern \(p\), then the
contribution of pairs of pattern \(p\) to the \(\alpha(\mathbf{Z})\)
vector is simply \(H_p^m \times h_p\). Thus our posterior update for
\(\alpha\) is simply
\(\alpha(\mathbf{Z}) = \alpha_0 + \sum_{p=1}^P H_p^m \times h_p\). Then,
we can easily calculate \(H_p^u\), the number of nonmatching record
pairs of agreement pattern \(p\), by subtracting the number of matching
pairs from the total present in the data; that is
\(H_p^u = H_p - H_p^m\). From this, we can update our \(\beta\)
parameter through
\(\beta(\mathbf{Z}) = \beta_0 + \sum_{p=1}^P H_p^u \times h_p\). Note
that these constitute \(P\) many summations over \(n_B\) quantities, and
thus avoid the \(n_A \times n_B\) summation from the original method.

Sadinle uses a prior for \(\mathbf{Z}\) that induces the a full
conditional for \(Z_j\) that strictly enforces one-to-one matching. In
particular, this sampler removes previously matches records from the set
of candidate records when sampling \(Z_j\), creating a dependency that
makes the sampler \emph{inherently serial}. By weakening the one-to-one
requirement, our full conditional for \(Z\) does not depend on the rest
of the \(\mathbf{Z}_{-j}\) vector, and thus can be computed in parallel.
More importantly, since only the agreement pattern of \(Z_j\) is used
for calculations within the Gibbs sampler, and not the particular record
label, we can conduct this sampling only at the level of the unique
agreement patterns. This boosts computation time far greater than
parallelization.

To do this, we calculate the Fellegi Sunter weight \(w_{h_p}\) for each
unique pattern, sample the agreement pattern between \(j\) and its
potential match, and then sample the record label uniformly among viable
records. More concretely, define \(h(Z_j)\) to be the agreement pattern
between \(j\) and its potential match, and say \(h(Z_j) = h_{P+1}\) when
\(Z_j = n_A + 1\). Then,

\[h\left(Z_j^{(s+1)}\right) | \mathbf{m}, \mathbf{u}, \mathbf{Z^{(s)}} \propto
\begin{cases} 
    w_{h_p}\times H_{p_j}  & p \leq P; \\
     n_A \frac{n_B - D + \beta_{\lambda}}{D + \alpha_{\lambda}} &   p = P + 1 \\
\end{cases}\]

@@

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by
running \texttt{BRL} on simulated pairs of files with differing levels
of errors and overlap. Comparing our method against \texttt{BRL} on
these same simulated datasets, we find that our method only has weakened
performance in the extreme scenario of very high errors and very high
overlap across files. We use the following measures of accuracy in our
analysis: although using only recall and precision is common in the
literature, we find it useful to have a single metric the gauge overall
performance, and thus also include F measure.

\[\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}\]
\[\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}\]
\[\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)\]

\begin{center}\includegraphics[width=29.17in]{../notes/figures/sadinle_sim_plot} \end{center}

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month, and and year of
birth, with 1 indicating an exact match on a particular field and 0
indicating a nonmatch. We simulate these data for different values of
\(n_A\) and \(n_B\), and compare the runtime of \texttt{parlr} against
\texttt{BRL}. We see that at low data size, \texttt{BRL} outperforms,
but that \texttt{parlr} is significantly faster at handling larger data.

In particular, runtime for \texttt{BRL} seems to grow quadratically
while runtime for \texttt{parlr} seems to grow linearly.

\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & m & u \\ 
  \hline
fname\_0 & 0.05 & 0.99 \\ 
  fname\_1 & 0.95 & 0.01 \\ 
  lname\_0 & 0.05 & 0.99 \\ 
  lname\_1 & 0.95 & 0.01 \\ 
  day\_0 & 0.05 & 0.97 \\ 
  day\_1 & 0.95 & 0.03 \\ 
  month\_0 & 0.05 & 0.92 \\ 
  month\_1 & 0.95 & 0.08 \\ 
  year\_0 & 0.05 & 0.93 \\ 
  year\_1 & 0.95 & 0.07 \\ 
   \hline
\end{tabular}
\end{table}

\begin{center}\includegraphics[width=23.6in]{../notes/figures/sadinle_speed_plot} \end{center}

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Shockingly, this
seems to be true. In the plot below, fixing \(n_B = 500\), we see linear
growth for the runtime under \texttt{BRL} as \(n_A\) increases, with
much more static runtime under \texttt{parlr}. The slight increases in
runtime that we do see are due primarily to the hashing step, which
again can be run in parallel for large data.

\begin{center}\includegraphics[width=29.17in]{../notes/figures/speed_plot_fixed_nB} \end{center}

We note here that \texttt{BRL} is coded in C, which makes for unfair
comparison against \texttt{parlr}, currently only built in R.
Additionally, although \texttt{parlr} is amenable to parallezation, this
simulation was run on a single core. Running \texttt{parlr} in C++ with
paralellization for the hashing step and sampling the matching status of
the record pairs should lead to even more drastic results.
